---
title: "K-mer Basics"
description: A brief review of k-mers and their relevant properties.
tutorial:
  id: "kmerize.basic"
  version: 0.9
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(kmerize)
library(ggbio)
```



## What Are K-mers?

K-mers are DNA sequences of length k. The can be generated by moving along a mother- or template DNA in steps of 1. For k = 1 - the shortest - this equals the set of nucleotides A, C, G, and T. Other well-established k-mers include dimers and trimers. Here we are interested in longer k-mers typically in the range from k > 3 to k-mers of lenght < 100 or also occasionally longer.

### What are they good for?

K-mers have found a variety of applications. Particularly interesting is their use as fingerprints at different levels of resolution. In metagenomics, for example, the mixed sample of DNA from a host of organsims is typically annotated with species labels by screening against a database of k-mer signatures.

## Main challenges

A major challenge is to decide on a minimal k-mer size that still is capable to support the envisioned application. The longer the k-mers become the more combinations are possible and so an index in a database will become very big or too big. Along with this, the time to create and consult the database will increase. While this can be remedied to a certain extent with parallel computing, it is still desirable and cheaper, to keep the k-mer size small.

Due to the many still unknown factors in genome biology it is yet not possible to calculate exactly how many distinct k-mers to expect for a certain k and a given species. In general we expect, with increasing k to have more unique k-mers. We also expect, that within a species the unqiue k-mers are limited.

In principle, it is possible to heuristically estimate the total number of k-mers, the distinct, and the unique k-mers by systematically scanning a genome for a certain k range. Experience has shown that it is convenient to only consider kmers of of uneven numbers. 

Here, the example of the virus phix174 (the first sequenced organism) shows the principle following a blogpost here(x). Another wording is, to evaluate the genomic response to changes in k.

## Genomic Response of Phi x 174 

This package contains the full sequende of PhiX174 from NCBI. 

```{r eval=FALSE}
library(kmerize)

```


```{r data, message=FALSE}
  # Getting a genome
  fa <-  system.file("testdata/phix174.fasta", 
                     package = "kmerize")
                                          
  phix_length <- kmr_length_fasta(fa)                   

  phix_length
  
  # Setting an output file in a temporary directory since this is for demo
  out_file <- file.path(tempdir(), "phix174")
  
  # here we create a sequence of k in steps of 2

  k <- as.integer(seq(3, 21, 2))
  
  phix_response <- kmr_response(fa, k)

```



## Genomic Response Table

The convenience method for the genomic response returns a summary table. This includes also the theoretical maximum number of k-mers (the k-mer universe) for each value of k.

The resulting table shows only values up to 13. This is because currently, it stops once the number of unique k-mers equals (exactly more than .9999) the distinct k-mers. This means, at and after this k value all k-mers have become unique.

```{r}

knitr::kable(phix_response)

```





## Genomic Response Graph

Plotting this table shows the following response curve:

```{r grplot, fig.width=7}


 p <- kmr_plot_response(phix_response, max_y = phix_length)

  p

```


Now the curves for the unique and distinct k-mers follow an S curve or power law curve with diminishing response in variability (distinct / unique) as k increases. At around 95% there seems to be a distinct land mark.

It could be an example of the 95/5 rule going back to Deming:
He argued that 95% of an organization’s performance depends on the quality of the system, and that only 5% is attributable to the actions of discrete individuals.
In analogy, one could expect that most of an organisms makeup is needed for basic house keeping and maybe 5% are interesting 'diversity'.


```{r, echo=FALSE}
lim <- 0.95

k_lim <- max(phix_response$total) * lim

k_lim
```


So, this cutoff value can be added to the graph as follows:


```{r, fig.width=7}
  p <- p +
      geom_hline(yintercept = k_lim, linetype = "dashed", color = "darkgreen", lwd = .6)
 p 

```

And the corresponding k value can be extracted with the following function:

## Convencience Graph Options

The plot function includes options to automatically show a k cutoff corresponding to a certain limit and also to set a maximum limit for y axis.

```{ plot3r}

```{r fig.width=7}
  p <-  kmr_plot_response(phix_response, max_y = phix_length, limit = lim)
 p
```
```

